TODO: title

### 2014-05-06 paper/main.tex
finished metrics subsection

%TODO: diversity config matches item-item config.
%TODO: review results
%TODO: outline conclusions
%TODO: write introduction
%TODO: write methodology
%TODO: re-write results
%TODO: write conclusions
%TODO: review extensively
\documentclass[letterpaper]{sig-alternate}

@@

  We seek to develop an understanding of how a range of different standard algorithms perform at for new users.
  Rather than picking advanced algorithms that are tuned for new user performance, we decided to first understand how representatives of three common types of algorithms perform.
  Therefore we will compare User-User, Item-Item, and Funk's SVD algorithms against two simple baseline algorithms.
  More information about these algorithms can be found in table \ref{tbl:algo}
  Code for our evaluation, using the Lenskit evaluation framework \cite{lenskit} is available at \url{https://bitbucket.org/kluver/coldstartrecommendation}.

  \begin{table}
    \centering
    \begin{tabular}{|p{6em}|p{18em}|}
      \hline
      Algorithm          & Algorithm description \\\hline
      ItemBaseline       & Item's average rating, with mild bayesian damping towards the global mean \cite{funk_netflix_2006}. \\\hline
      UserItem\-Baseline & ItemBasline adjusted by the user's average offset from the ItemBaseline. Mild mean damping is applied\cite{funk_netflix_2006}. \\\hline
      User-User          & User based nearest neighbor collaborative filtering \cite{resnick1994grouplens} with a neighborhood size of 30 and ratings normalized by subtracting the UserItemBaseline score. \\\hline
      Item-Item          & Item based nearest neighbor collaborative filtering \cite{sarwar2001item} with a neighborhood size of 30 and ratings normalized by subtracting the ItemBaseline score.   \\\hline
      SVD                & Simon Funk's SVD based collaborative filtering approach \cite{funk_netflix_2006} with 30 features and 150 training iterations per feature. \\\hline
    \end{tabular}
    \caption{Summary of algorithms.}
    \label{tbl:algo}
  \end{table}

  % evaluation and code to run these algortihms available at.
  % citations and lenskit info
  Each algorithm was configured and tuned similarly to Ekstrand's 2012 evaluation \cite{ekstrand2012recommenders}.
  Since this evaluation was on a slightly different dataset (MovieLens 10M), we performed minor tuning from these parameters.
  Most notable, we found that the previous damping parameter used in the baselines was too large.
  We found that a value of 5 gave much better results for new user recommendations.
  During tuning we found that, among the configurations that performed well, our key results are consistent.
  %TODO: carefully run this wording by joe.

  \subsection*{Metrics}
  We will discuss a total of eleven metrics.
%%%%%%%%%%%%%%%%%%%%%%
  Some of these metrics will prove to be redundant, and future work should be able to use a much smaller list.
  The metrics are listed in table \ref{tbl:metrics}.
  \begin{table}[ht!]
    \centering
    \begin{tabular}{|p{7em}|p{16em}|}
      \hline
      \multicolumn{2}{|c|}{{\bf Accuracy Metrics}} \\\hline
      RMSE                   & Standard prediction accuracy metric. \\\hline
      \hline
      \multicolumn{2}{|c|}{{\bf Recommendation Metrics}} \\\hline
      nDCG                   & Standard ranking quality evaluation.\\\hline
      Precision@20           & Count of recommended items in the test set and rated 4.0 or higher.\\\hline
      MAP@20                 & Average of precision at each rank one through twenty.\\\hline
      Fallout@20             & Count of recommended items in the test set and rated 2.0 or lower.\\\hline
      SeenItems@20           & Count of recommended items in the test set.\\\hline
      MeanRating@20          & Average rating of recommended items that are in the test set.\\\hline
      RMSE@20                & RMSE computed only over those items that were recommended.\\\hline
      \hline
      \multicolumn{2}{|c|}{{\bf Other Metrics}} \\\hline
      Average\-Popularity@20 & The average number of users in the training set that have seen recommended items.\\\hline
      AILS@20                & The average pairwise similarity between recommended items. \\\hline
      Spread@20              & The Shannon's entropy of the distribution of recommended items for users in the test set.\\\hline
    \end{tabular}
    \caption{TODO:}
    \label{tbl:metrics}
  \end{table}

  All metrics evaluating recommendations (all but RMSE and nDCG) are based on the top 20 predicted items for the user, excluding items in the training set.
  The choice of 20 items for the evaluations was arbitrary, testing with other values showed similar results, except on one metric.
  This issue will be addressed later in the results section.

  Prediction accuracy, as measured by RMSE or MAE is one of the classic techniques in recommended system evaluation.
  The idea behind these evaluations is to measure the accuracy of predictions on withheld recommendations.
  If the algorithm can successfully predict the withheld ratings, then we expect it will accurately predict future ratings as well.
  Because users may build trust in the system by looking at the predictions of items, we are interested in knowing which algorithms are most accurate, and which show the most notable change in accuracy as ratings are entered.

  More recently, the research community has shifted away from accuracy metrics in favor of metrics that evaluate a recommender system's ability to rank or recommend.
  These algorithms try to evaluate how useful the recommendations provided by the system are to the user.
  This is commonly done by looking either at how withheld items are ranked with metrics such as nDCG, or by looking at if good or bad withheld items tend to be recommended positively by the recommender.

  Prior work has shown that users are sensitive to properties of a recommender other than simply how well it performs.
  These properties, such as the diversity of the output recommendations, can effect how the user interacts with the recommender and drastically change the user experience \cite{zieglerDiversity, martijnDiversity}.
  Unfortunately, there is no way to know all such interesting metrics
  We will look at three such metrics, the popularity, diversity, and spread of the recommendations.

  The popularity of the items recommended to a user can have an impact on the user's satisfaction.
  If the user is only recommended items that are very popular, then the recommender is likely failing to make novel and interesting recommendations.
  In this case the user may feel that the recommender is not helpful in finding new and interesting things to watch.
  However, if a new user is only recommended items that are not well known, it is unlikely that the user will have heard of the items recommended.
  This could cause the user to not trust the system, as they have no way of comparing the recommendations against their personal knowledge.

  Past work \cite{zieglerDiversity, martijnDiversity} has shown that users tend to prefer recommendation lists that are more diverse.
  The belief is that by providing more diverse lists we are giving the user a larger range of items to choose from.
  This can make choosing an item easier as the user is less likely to have to compare very similar options.

  Finally, we expect that algorithms with a good understand of its users will be able to recommend different items to different users.
  To measure this we will look at how well the recommender spreads its attention of many items.
  We will do this by computing the Shannon's entropy of the items recommended to each user.
  If the recommender system recommends the same items to each user, then the entropy will be low.
  Conversely, if the recommender recommends different items to each user, then the entropy will high.
  In practice, a recommender may focus on both a shared core of popular or well reviewed items, and more personal items, which would lead to middle values.

  Based on this list of possible metrics, we will clarify our research questions with three sub questions:

  % Provide computation details for AILS@N
  Most of our metrics are standard techniques, or easily commutable, therefore we will only comment on the AILS, and spread metrics.
%%%%%%%%%%%%%%%%%%%%%%%%%
  The AILS@20 metric is based on the ILS metric introduced by ziegler et al \cite{zieglerDiversity}.
  AILS@N is simply a rescaling of ILS to be scale free with regards to N, this was done to make the metric easier to interpret, by putting it on the same scale as similarity values.
  We used the same item similarity metric for AILS@20 as we used in the Item-Item recommender.

  % Provide computation details for spread
  The spread metric is computed by computing recommendations for each test user.
%%%%%%%%%%%%%%%%%%%%%%%%%%  
  Let $C(I)$ be the count of how many times item $I$ showed up in the recommendations.
  Let $P(I) = C(I) / \sum_I C(I)$ be the probability that a random recommendation is item $I$.
  Finally we define spread as the Shannon's entropy of $P(I)$, $Spread = -\sum_I P(I) \log(P(I))$.
%%%%%%%%%%%%%%%%%%%%%%%%%%

  % provide expectations.
  We expect that the item baseline algorithm will have approximately constant performance on all metrics.
  Since the Item Baseline is such a simple algorithm we expect that any reasonable algorithm should be able to do better on it on the accuracy on recommendation metrics.
  Furthermore, we expect the improvement over the item baseline to be gradual, with larger improvements only happening with more data.

  Many of the algorithms are configured with 
  We expect that our intelligent algorithms, SVD, Item-Item, and User-User should be able to beat the item user baseline after

>>>>

%TODO: review results
%TODO: outline conclusions
%TODO: write introduction
%TODO: write methodology
%TODO: re-write results
%TODO: write conclusions
%TODO: review extensively
\documentclass[letterpaper]{sig-alternate}

@@

  We seek to develop an understanding of how a range of different standard algorithms perform at for new users.
  Rather than picking advanced algorithms that are tuned for new user performance, we decided to first understand how representatives of three common types of algorithms perform.
  Therefore we will compare User-User, Item-Item, and Simon Funk's SVD algorithms against two simple baseline algorithms.
  More information about these algorithms can be found in table \ref{tbl:algo}
  Code for our evaluation, using the Lenskit evaluation framework \cite{lenskit} is available at \url{https://bitbucket.org/kluver/coldstartrecommendation}.

  \begin{table}
    \centering
    \begin{tabular}{|p{6em}|p{18em}|}
      \hline
      Algorithm          & Algorithm description \\\hline
      ItemBaseline       & Item's average rating, with mild bayesian damping towards the global mean \cite{funk_netflix_2006}. \\\hline
      UserItem\-Baseline & ItemBasline adjusted by the user's average offset from the ItemBaseline. Mild mean damping is applied \cite{funk_netflix_2006}. \\\hline
      User-User          & User based nearest neighbor collaborative filtering \cite{resnick1994grouplens} with a neighborhood size of 30 and ratings normalized by subtracting the UserItemBaseline score. \\\hline
      Item-Item          & Item based nearest neighbor collaborative filtering \cite{sarwar2001item} with a neighborhood size of 30 and ratings normalized by subtracting the ItemBaseline score.   \\\hline
      SVD                & Simon Funk's SVD based collaborative filtering approach \cite{funk_netflix_2006} with 30 features and 150 training iterations per feature. \\\hline
    \end{tabular}
    \caption{Summary of algorithms.}
    \label{tbl:algo}
  \end{table}

  % evaluation and code to run these algortihms available at.
  % citations and lenskit info
  Each algorithm was configured and tuned similarly to Ekstrand's 2012 evaluation \cite{ekstrand2012recommenders}.
  Since this evaluation was on a different dataset (MovieLens 10M vs. MovieLens 1M), we performed minor tuning from these parameters.
  Most notable, we found that the previous damping parameter used for the baselines was too large.
  We found that a value of 5 gave much better results for new user recommendation.
  During tuning we found that our key results are consistent among those configurations that perform well.
  %TODO: carefully run this wording by joe.

  \subsection*{Metrics}
  We will discuss a total of eleven metrics.
  Some of these metrics will prove to be redundant, and future work should be able to use a much smaller list of metrics.
  The metrics are listed in table \ref{tbl:metrics}.
  \begin{table}[ht!]
    \centering
    \begin{tabular}{|p{7em}|p{16em}|}
      \hline
      \multicolumn{2}{|c|}{{\bf Accuracy Metrics}} \\\hline
      RMSE                   & Standard prediction accuracy metric. \\\hline
      \hline
      \multicolumn{2}{|c|}{{\bf Recommendation Metrics}} \\\hline
      nDCG                   & Standard ranking quality evaluation.\\\hline
      Precision@20           & Count of recommended items in the test set and rated 4.0 or higher.\\\hline
      MAP@20                 & Average of precision at each rank one through twenty.\\\hline
      Fallout@20             & Count of recommended items in the test set and rated 2.0 or lower.\\\hline
      SeenItems@20           & Count of recommended items in the test set.\\\hline
      MeanRating@20          & Average rating of recommended items that are in the test set.\\\hline
      RMSE@20                & RMSE computed only over those items that were recommended.\\\hline
      \hline
      \multicolumn{2}{|c|}{{\bf Other Metrics}} \\\hline
      Average\-Popularity@20 & The average number of users in the training set that have seen recommended items.\\\hline
      AILS@20                & The average pairwise similarity between recommended items. \\\hline
      Spread@20              & The Shannon's entropy of the distribution of recommended items for users in the test set.\\\hline
    \end{tabular}
    \caption{TODO:}
    \label{tbl:metrics}
  \end{table}

  All metrics evaluating recommendations (all but RMSE and nDCG) are based on the top 20 highest predicted items for the user, excluding items in the training set.
  The choice of 20 items for the evaluations was arbitrary, testing with other values showed similar results, except on one metric.
  This issue will be addressed later in the results section.

  Most of our metrics are standard techniques, so we will only address the latter 6.
  As we will discuss later, we found Precision, MAP, and fallout to not be useful for this evaluation, therefore to help understand the quality of our recommendations, we include SeenItems@20, MeanRating@20, and RMSE@20.
  SeemItems@20 tells us how frequently the user is to have previously seen a recommended item.
  This is useful for two reasons.
  First this provides important context for the next two metrics, as this is how many items those metrics are often averaged over.
  Secondly, if too many items are known to the user, the user might feel that the recommendations are not interesting.
  Alternatively, if too few of the items are known to the user, the user may have trouble evaluating the recommendations, which could lead the user to leave the system.

  MeanRatings@20, and RMSE@20 both try to establish how much the user likes the recommendations.
  MeanRatings@20 tells us directly how much the user tends to like recommended movies.
  RMSE@20 accompanies this and lets us know if the recommended items are likely to be mispredicted, which can explain why users may or may not like their recommendations.

  The average popularity metric is a crude metric of how novel the recommendations are to the user \cite{zieglerDiversity}.
  Broadly speaking we expect users will prefer recommendation lists containing more novel (less popular) items.
  However, if the recommended items are too unpopular we risk that the user will not have heard of the items, which could make the user doubt the recommendations, hurting the user experiance.

  Past work has shown that users like their recommendations to be more diverse \cite{zieglerDiversity, martijnDiversity}.
  The belief is that by providing more diverse lists we are giving the user a larger range of items to choose from.
  This can make choosing an item easier as the user is less likely to have to compare very similar options.
  We will measure diversity with the AILS@20 metric.

  The AILS@20 metric is based on the ILS metric introduced by ziegler et al \cite{zieglerDiversity}.
  AILS@N is simply a rescaling of ILS to be scale free with regards to N, this was done to make the metric easier to interpret, by putting it on the same scale as similarity values.
  We used the same item similarity metric for AILS@20 as we used in the Item-Item recommender.

  The Spread@20 metric is designed to measure how well the recommender spreads its attention across many items.
  We expect that algorithms with a good understanding of its users will be able to recommend different items to different users.
  Spread is computed by computing recommendations for each test user.
  Let $C(I)$ be the count of how many times item $I$ showed up in the recommendations.
  Let $P(I) = C(I) / \sum_I C(I)$ be the probability that a random recommendation is item $I$.
  Finally we define spread as the Shannon's entropy of $P(I)$, $Spread = -\sum_I P(I) \log(P(I))$.
  When this is large we know that the recommender is recommending many items with a relatively even distribution.
  When this is low, the recommender is focusing on a small set of items which it recommends to every user.

  % provide expectations.
  We expect that the item baseline algorithm will have approximately constant performance on all metrics.
  Since the Item Baseline is such a simple algorithm we expect that any reasonable algorithm should be able to do better on it on the accuracy and recommendation metrics.
  Furthermore, we expect the improvement over the item baseline to be gradual, with larger improvements only happening with more data.

  The user-item baseline should only differ from the item baseline for metrics that use the prediction value.
  For all other metrics these two should provide the same responses.
  As the user-item baseline is still quite a simple baseline, we expect that each algorithm will beat this baseline as well.
  That said, for small rating values its possible that the improvements over this baseline will be very minimal.

  We expect that the baseline algorithms will provide the most popular recommendations, with the least diversity, and the least spread.
  We expect that any personalized algorithm should be able to make more nuanced recommendations than simply recommending the best rated items, which should translate as gains on these metrics.
